How not to constructivize cohomology

### 2018-10-05 slides-pssl104.tex
More work on slides

\newcommand{\pointthis}[3]{%
  \tikz[remember picture,baseline]{
    \node[anchor=base,inner sep=0,outer sep=0] (#2) {#2};
    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout relative pointer={(0.3cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.1cm)$) {#3};
  }%
}

@@

\section{Sheaf cohomology}

@@

          [ $\chi(X; \O_X) = 1 - \operatorname{genus}_X$, $(C \cdot C') = \chi(\O_C \otimes_{\O_X}^{\mathbb{L}} \O_C')$ ]
          \\[0.6em]
    \item Let~$A$ be an abelian group. Let~$X$ be semi-locally contractible.
          Then $H^n(X; \underline{A}) = H_{\text{sing}}^n(X; A)$.
          \\[0.6em]
    \item Let~$f : X \to Y$ be continuous. Then there is a spectral sequence
          $H^i(Y; R^jf_*(E)) \Rightarrow H^{i+j}(X; E)$.
  \end{itemize}
\end{frame}
%}

\section{Constructive mathematics}

@@

  \centering\includegraphics[width=0.5\textwidth]{andrej-bauer-five-stages} \\
  \scriptsize\href{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej Bauer, talk at the IAS}\par
  \small
  \bigskip

@@

      \footnotesize
      \mbox{``Every map~$\NN \to \NN$ is computable.''}
      \mbox{``Every map~$\RR \to \RR$ is continuous.''}
      \mbox{``Every map~$\affl \to \affl$ is polynomial.''}
      \mbox{``Heyting Arithmetic has exactly one model.''}
      \mbox{``The subsets of~$\{\heartsuit\}$ form a proper class.''}
      \mbox{``There is an injection~$\RR \to \NN$.''}
    \end{column}

@@

      \footnotesize
      \mbox{program extraction} \\
      \mbox{synthetic differential geometry} \\
      \mbox{synthetic algebraic geometry} \\
      \mbox{synthetic domain theory} \\
      \mbox{new reduction techniques in algebra} \\
      \mbox{Bohr topos for quantum mechanics}
    \end{column}
  \end{columns}
\end{frame}

@@

  \only<1>{
    \centering
    \rotatebox{90}{\tiny\scalebox{0.8}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
    \includegraphics[width=0.7\textwidth]{external-internal-small}
    \par
  }
  \pause

@@

\begin{frame}{On injective objects}
  \begin{block}{}
    A set~$M$ is \hil{injective} iff for any injection~$A \to B$,
    any map~$A \to M$ extends to a map on~$B$.
  \end{block}
  \vspace*{-1em}

@@

  \begin{block}{}
    \justifying
    A sheaf of modules~$M$ is \hil{injective} iff for any linear monomorphism~$A \to B$,
    any linear morphism~$A \to M$ extends to a linear morphism on~$B$.
  \end{block}
  \vspace*{-1em}

@@

  \begin{itemize}
    \small
    \item Injective sheaves of sets and injectives sheaves of modules are flabby.
    \item Every sheaf embeds into a flabby sheaf.
    \item Assuming the law of excluded middle, any sheaf of modules over a
    topological space embeds into a flabby sheaf of modules.
    \item Assuming Zorn's lemma, flabby sheaves of modules are \hil{acyclic for the global sections functor}.
    \item Assuming \bad{??}, sheaf cohomology and higher direct images can be computed using \hil{flabby resolutions}.
  \end{itemize}

  \begin{block}{}
    A set~$M$ is \hil{flabby} iff for any subterminal~$K \subseteq M$, there is
    an element~$x \in M$ such that~$K \subseteq \{x\}$.
  \end{block}
  \vspace*{-1em}

@@

% XXX: Results about Eff
% XXX: Summary slide

>>>>

\newcommand{\pointthis}[3]{%
  \tikz[remember picture,baseline]{
    \node[anchor=base,inner sep=0,outer sep=0,color=blue!90] (#2) {#2};
    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout
    relative pointer={(0.0cm,-0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,0.7cm)$) {#3};
  }%
}

\newcommand{\pointthisbelow}[3]{%
  \tikz[remember picture,baseline]{
    \node[anchor=base,inner sep=0,outer sep=0,color=blue!90] (#2) {#2};
    \node[visible on=#1,overlay,rectangle callout,rounded corners,callout
    relative pointer={(0.0cm,0.5cm)},fill=blue!20] at ($(#2.north)+(-0.1cm,-1.0cm)$) {#3};
  }%
}

@@

\section[Cohomology]{Sheaf cohomology}

@@

          [ $\chi(X; \O_X) = 1 - \operatorname{genus}_X$,
          $(C \cdot C') = \chi(\O_C \otimes_{\O_X}^{\mathbb{L}} \O_C')$, \ldots ]
          \\[0.6em]
    \item Let~$A$ be an abelian group. Let~$X$ be semi-locally contractible.
          Then $H^n(X; \underline{A}) = H_{\text{sing}}^n(X; A)$.
          \\[0.6em]
    \item Let~$f : X \to B$ be continuous. Then there is a spectral sequence
          $H^i(B; R^jf_*(E)) \Longrightarrow H^{i+j}(X; E)$.
  \end{itemize}
\end{frame}
%}

\section[Constructivism]{Constructive mathematics}

@@

  \centering\includegraphics[width=0.5\textwidth]{andrej-bauer-five-stages} \\
  \scriptsize\href{https://video.ias.edu/members/1213/0318-AndrejBauer}{Andrej Bauer at an IAS talk}\par
  \small
  \bigskip

@@

      \footnotesize
      \mbox{``Every map~$\NN \to \NN$ is computable.''}
      \mbox{``Every map~$\RR \to \RR$ is continuous.''}
      \mbox{``Every map~$\affl \to \affl$ is polynomial.''}
      \mbox{``Heyting Arithmetic has exactly one model.''}
      \mbox{``The subsets of~$\{\heartsuit\}$ form a proper class.''}
      \mbox{``There is an injection~$\RR \to \NN$.''} \\
      \rotatebox{90}{\dots}
    \end{column}

@@

      \footnotesize
      \mbox{program extraction} \\
      \mbox{synthetic differential geometry} \\
      \mbox{synthetic algebraic geometry} \\
      \mbox{synthetic domain theory} \\
      \mbox{new reduction techniques in algebra} \\
      \mbox{Bohr topos for quantum mechanics} \\
      \rotatebox{90}{\dots}
    \end{column}
  \end{columns}
\end{frame}

@@

  \only<1>{
    \vspace*{0.5em}
    \centering
    \rotatebox{90}{\tiny\scalebox{0.8}{Illustration: Carina Willbold}}\hspace{-0.05cm}%
    \includegraphics[width=0.7\textwidth]{external-internal-small}
    \par
  }
  \pause

@@

\begin{frame}{Internalizing higher direct images}
  \begin{block}{}
    A set~$M$ is \hil{injective} iff for any injection~$A \to B$,
    any map~$A \to M$ extends to a map on~$B$.
  \end{block}
  \vspace*{-1em}

@@

  \begin{block}{}
    \justifying
    A sheaf of modules~$M$ is \hil{injective} iff for any linear monomorphism
    \mbox{$A \!\!\to\!\! B$,
    any linear morphism~$A \!\!\to\!\! M$ extends to a linear morphism on~$B$.}
  \end{block}
  \vspace*{-1em}

@@

  \begin{itemize}
    \small\justifying
    %\item Injective sheaves of sets and injectives sheaves of modules are flabby.
    \item Every sheaf embeds into a flabby sheaf.
    \item Assuming the law of excluded middle, any sheaf of modules over a
    topological space embeds into a flabby sheaf of modules.
    \item Assuming Zorn's lemma, flabby sheaves of modules are \hil{acyclic for the global sections functor}.
    Hence, assuming \bad{??}, sheaf cohomology and higher direct images can be computed using \hil{flabby resolutions}.
  \end{itemize}

  \bigskip
  \bigskip

  \begin{block}{}
    A set~$M$ is \hil{flabby} iff any \pointthis{<1>}{subterminal}{$\forall x,y
    \in K\_ x = y$}~$K \subseteq M$ is a \pointthis{<1>}{subsingleton}{$\exists
    a \in M\_ \forall x \in K\_ x = a$}.
  \end{block}
  \vspace*{-1em}

@@

\begin{frame}{Flabbiness as an organizing principle}
  \begin{block}{}
    A set~$M$ is \hil{flabby} iff any \pointthisbelow{<1>}{subterminal}{$\forall x,y
    \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
    a \in M\_ \forall x \in K\_ x = a$}.
  \end{block}
  \bigskip
  \bigskip
  \justifying

  \textbf{Proposition.} Let~$E$ be a sheaf over a locale~$X$. Then~$E$ is flabby
  iff it is a flabby set from the point of view of~$\Sh(X)$.

  \textbf{Proof.} Routine/Zorn's lemma.

  \textbf{Proposition.} Let~$M$ be a sheaf of modules over a locale~$X$.
  Then~$M$ is injective iff it is injective from the point of view of~$\Sh(X)$.

  \textbf{Proof.} (Only ``$\Leftarrow$''.) Let~$i : A \to B$ be a linear
  monomorphism. Let~$f : A \to M$ be a linear morphism. Then verify, internally, that
  the set~$E \defeq \{ \bar f : B \to M \,|\, \bar f \circ i = f \}$ is flabby.
%  Let~$K \subseteq E$ be a subterminal.
%  We consider the injectivity diagram
%  \[ \xymatrix{
%    i[A] + B' \ar@{^{(}->}[r]\ar[d]_g & B \ar@{-->}@/^/[ld] \\
%    I
%  } \]
%  where~$B'$ is the submodule~$\{ s \in B \,|\, \text{$s = 0$ or $K$ is
%  inhabited} \} \subseteq B$ and the solid vertical arrow~$g$ is defined in the following
%  way: Let~$s \in i[A] + B'$. Then~$s = i(a) + t$ for an element~$a \in A$ and an
%  element~$t \in B'$. Since~$t \in B'$, $t = 0$ or~$K$ is inhabited. If~$t = 0$,
%  we set~$g(s) \defeq f(a)$. If~$K$ is inhabited, we set~$g(s) \defeq f(a) +
%  \bar{f}(s)$, where~$\bar{f}$ is any element of~$K$. These prescriptions
%  determine a well-defined map.
\end{frame}

\section[In the effective topos]{Flabbiness in the effective topos}

\begin{frame}{Flabbiness in the effective topos}
  \begin{block}{}
    A set~$M$ is \hil{flabby} iff any \pointthisbelow{<1>}{subterminal}{$\forall x,y
    \in K\_ x = y$}~$K \subseteq M$ is a \pointthisbelow{<1>}{subsingleton}{$\exists
    a \in M\_ \forall x \in K\_ x = a$}.
  \end{block}
  \bigskip
  \bigskip
  \bigskip

  \justifying
  \textbf{Proposition.}
  Let~$X$ be an effective object in the effective topos. Then
  \[ \text{``If~$X$ is flabby, any endomap on~$X$ has a fixed point.''} \]
  from the point of view of the effective topos.
  \medskip

  \textbf{Proposition.}
  Assuming the law of excluded middle, any~$\neg\neg$-separated
  module in the effective topos can be embedded into a flabby module.
  \medskip

  \textbf{Question.} Are there enough flabby modules in the effective
  topos?
\end{frame}

% XXX: ctcl.pdf, realizability book
% XXX: Harting

\begin{frame}{State of affairs}
  \centering

  \textcolor{blue!90}{\Huge :-)}

  The existence of enough flabby modules is constructively neutral. \\
  Higher direct images can be understood as internal sheaf cohomology.

  \bigskip

  \bad{\Huge \scalebox{2}{:-(}}

  Flabby sheaves can fail to be acyclic, constructively. \\
  There is still no general constructive framework for sheaf cohomology.
  Even though:

  \begin{itemize}
    \item Basic homological algebra is entirely constructive.
    \item There are algorithms for computing cohomology [Barakat XXX].
    \item Čech methods work constructively, even in a synthetic context.
  \end{itemize}
\end{frame}
\addtocounter{framenumber}{-1}
